Physics-based protein structure prediction and design using the confinement method

### 2012-10-01 Paper/paper.tex
second draft arijit

However, the theoretical calculation of conformational free energy change become difficult if the
states of interest are very different. In such scenario timescale involved in such tranformation may
be beyond timescale of direct classical molecular dynamics simulation.  Generally, in order to
calculate free energy a pathway or reaction coordinate is created with the intermediate structures.
Method like umbrella sampling, thermodynamic integration along with classical MD can suffer from
overlap problem and require large computational effort. The idea of reaction coordinate became more
complicated in case of protein folding.  As free energy landscape of the protein folding is rough,
it can be trapped in a energy well during its visit in the conformational space. Even with the
advanced method like replica exchange molecular dynamics the three dimensional structure may be
trapped in a local minima for a considerable time. Due to various reasons it can also be misfolded
and trapped in a local energy minima. And one can mistakenly identify the misfolded state as the
native state (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

Calculation of free energy has been successfully attempted by a number of groups (give examples). In
recent years methods like Refrence system method, Decativated Morphing, Orthogonal Space Random
Work, Confinement Method have came out for calculation of free energy method. Some of the groups
relies on one of the great strength of free energy calculation, that it is a state function and thus
it is independent of path.

In this work we have applied the confinement method which was originally devoled by Tyka et. al. and
Cecchini et. al.  This method relies on a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraint to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Then we have applied this method to larger proteins for the
first time. The free energy of conformation of pair of proteins with similar sequence but completely
different fold was calculated. Encouraged by the result, we applied the method of different targets
of CASP9 competition.

@@

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from
protein G, known as BHP[ref]. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations[ref] shows that bhp1 is the more favorable
configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value 1.7 kcal/mol, which
is in good agreement with the equilibrium simulations and with previous calculations using the
confinement method[ref] (see Supporting Information for further details).

@@

Previous applications of the confiment method have focused on relatively short peptides, up to X
residues in length \Justin{Arijit: fill in X}. In this work we apply the confinement method to
larger proteins. One of the  applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9 and 10, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimental structure (once it was available from the PDB). As
Table~\ref{table:casp_control} shows, in X of Y cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. Although
differentiating between experimentally determined structures and computer generated predictions is
not a stringent test of a scoring method---for example, examining the residue-residue packing can
easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
"sanity check".

@@

\begin{center}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
\end{tabular}
\end{center}

@@

The fact such small changes in sequence can lead to such dramatic changes in structure is rather
remarkable. Accurately predicting the structural preferences of these structures presents a serious
challenge for computational methods [ref].

@@

The results of these calculations are presented in Figure~\ref{fig:orban}. The confinement method
identifies the correct structure for all six sequences. The free energy differences range from
around 2.9 to 5.7 kcal/mol, which is consistent with small changes in sequences and with estimates
made by Orban and co-workers [ref].

\begin{figure}
\label{fig:orban}

\caption{The confinement method correctly predicts the structural preferences of six chameleon
sequences. (A) The six sequences used in this study. (B) Each sequence adopts either a Protein
G-like fold (denoted GB) or a three-helix bundle fold (denoted GA). The relative free energies of
the two folds are reported for each sequence.}

\Justin{Arijit: we need to make this figure, which we've discussed before and move the data from the
following table into the figure.}
\end{figure}

@@

The native structure is a  67 residue NMR solved structure. We first reviewed all the five models
submitted by the group "BAKER-ROSETTASERVER", which was the best predictor group for this particular
target.  Among them we choose only the three structures, model 1, 3 and 5 to calculate the free
energy and subsequently identify the most stable structure among them. The rest of the two
structures are close to either of the selected structure and thus they are ommited from our
calculation. According to the GDT\_TS score, the model 1 was picked up correctly from the set of
structure generated . However, the order of model 3 and 5 was not correct as can be seen from the
Table 1. It will be interesting to know whether confinement method can order the strcutures in terms
of free energy scale. The result is presented in Table~\protect\ref{tab:T0559}.

@@

Target T0560 is a 74 amino acid residue target from CASP9. However residues 3-66 were kept for final
result.  For this target we have compared the free energy of the model 2 and model 4 submitted by
the group "Splicer" along with the crystal structure.  The main difference between the model 2 and 4
is the orientation of the ....  The final result is presented in Table~\protect\ref{tab:T0560}.

@@

Target T0538 was a 54 residue target. The native structure was generated using NMR. The best model
that is closest to the native structure was generated by the group PconsR. However, their model 3
was better than their model 1. This structure was quite close to the NMR structure and had GDT\_TS
value of 96.23. We have calculated the free energy between this structure and the best structure
submitted by the group "Shell", "FOLDIT" and "BAKER-ROSETTASERVER". The calculated free energy value
is presented in Table~\protect\ref{tab:T0538}.

@@

Target T0540 was a 90 residue target. We have taken the best models from the group "LTB" which was
the best performing group for this target and also from the group Mufold and calculated the free
energy difference between these models with that of the native crystal structure.  As expected, the
crystal structure has the lowest free energy value, followed by Model 1 with GDT\_TS score of
$69.72$ and model 2 with GDT\_TS value of $53.789$. The final result is presented in
Table~\protect\ref{tab:T0540}.

@@

This is a 65 amino acid residue target with residues 6-63 were kept for final analysis. We have
calculated the free energy difference between all the five models submitted by the group
"MUFOLD-MD", which was one of the best performed group for this target.  The result is presented in
Table~\protect\ref{tab:T0531}

@@

The best performing group was "Baker" for this target. We have carried out our calculation with the model first, second and fifth for the
final calculation. The rest two model had closer GDT\_TS value compared to the previously mentioned models and thus they were omitted from
furthe calculation. The result is presented in Table~\protect\ref{tab:T0605}. (This is a case where we can demand the confinement
method is better than GDT\_TS).

@@

\begin{figure}
\begin{center}
\includegraphics[width=12cm,height=10cm]{G88.pdf}
\end{center}
\caption{Two protein with 88 \% similar sequence but different folds}
\label{fig:G88}
\end{figure}

@@

\begin{table}
\caption{The value of free energy and GDT\_TS for the CASP9 target T0559}
\label{tab:T0559}
\begin{center}
\begin{tabular}{l l l}\hline
Model   &     $\Delta G$ (Kcal/Mol) &  GDT\_TS \\ \hline
Model 1 &     0.000            &  94.03    \\
Model 3 &     $14.484 \pm 0.633$ &  48.13    \\
Model 5 &     $4.526  \pm 0.384$ &  82.46    \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}
\caption{The Free energy and GDT\_TS values of the CASP9 Target T0538}
\label{tab:T0538}
\begin{center}
\begin{tabular}{l l l l}\hline
Model   &     $\Delta G$ (Kcal/Mol) &  GDT\_TS & Group Submitted \\ \hline
Model 1 &     0.000              &  96.23    & PconsR  \\
Model 2 &     $15.891 \pm 0.475$ &  90.09  & Shell  \\
Model 3 &     $23.243 \pm 0.393$ &  86.32  & FOLDIT  \\
Model 4 &     $26.002 \pm 0.756$ &  80.66  & BAKER-ROSETTASERVER \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}
\caption{The free energy and the GDT\_TS value of the CASP9 target T0560.}
\label{tab:T0560}
\begin{center}
\begin{tabular}{l l l}\hline
Model   &     $\Delta G$ (Kcal/Mol) &  GDT\_TS \\ \hline
Crystal &     0.000              & 100.00    \\
Model 2 &     $22.153 \pm 0.491$ &  94.14    \\
Model 4 &     $27.428  \pm 0.477$ &  76.56    \\ \hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=12cm,height=10cm]{T0540.pdf}
\end{center}
\caption{T0540}
\label{fig:}
\end{figure}

\begin{table}
\caption{The free energy and the GDT\_TS value of the CASP9 target T0540.}
\label{tab:T0540}
\begin{center}
\begin{tabular}{l l l l}\hline
Model   &     $\Delta G$ (Kcal/Mol) &  GDT\_TS & Group Submitted \\ \hline
Crystal &     0.000              &  100.00   & -      \\
Model 1 &     $22.003 \pm 0.575$ &  69.72    & LTB    \\
Model 2 &     $53.786 \pm 0.553$ &  63.89    & Mufold \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}
\caption{The free energy and the GDT\_TS value of the CASP9 target T0531.}
\label{tab:T0531}
\begin{center}
\begin{tabular}{l l l}\hline
Model   &     $\Delta G$ (Kcal/Mol) &  GDT\_TS \\ \hline
Crystal &     0.000              & 100.00    \\
Model 1 &     $12.496 \pm 0.704$ &  43.53    \\
Model 2 &     $27.374 \pm 0.691$ &  34.48    \\
Model 3 &     $13.914 \pm 0.673$ &  34.48    \\
Model 4 &     $53.055 \pm 0.684$ &  28.02    \\
Model 5 &     $29.508 \pm 0.698$ &  25.86   \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}
\caption{The free energy and the GDT\_TS value of the CASP9 target T0605.}
\label{tab:T0605}
\begin{center}
\begin{tabular}{l l l}\hline
Model   &     $\Delta G$ (Kcal/Mol) &  GDT\_TS \\ \hline
Model 1 &     $0.000$ &  $97.96$    \\
Model 2 &     $1.689 \pm ?$ &  $89.29$    \\
Model 5 &     $16.064 \pm ?$ & $93.88$   \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\bibliographystyle{achemso}
\bibliography{paper}

Ytreberg, F.; Zuckerman, D. Simple estimation of absolute free energies for biomolecules. J. Chem. Phys. 2006, 124, 104105.

Park, S.; Lau, A.; Roux, B. Computing conformational free energy by deactivated morphing. J. Chem. Phys. 2008, 129, 134102

Zheng, L.; Chen, M.; Yang, W. Random walk in orthogonal space to achieve efficient free-energy simulation of complex systems, Proc. Natl. Acad. Sci. 2008, 105 (51), 20227.

Tyka, M.; Clarke, A.; Sessions, R. An Efficient, Path-Independent Method for Free-Energy Calculations. J.Phys.Chem. B 2006, 110, 17212–17220.

>>>>

However, the theoretical calculation of conformational free energy change become difficult if the
states of interest are very different. In such scenario timescale involved in such tranformation may
be beyond timescale of direct classical molecular dynamics simulation.  Generally, in order to
calculate free energy a pathway or reaction coordinate is created with the intermediate structures.
Method like umbrella sampling, thermodynamic integration along with classical MD can suffer from
overlap problem and require large computational effort. The idea of reaction coordinate became more
complicated in case of protein folding.  As free energy landscape of the protein folding is rough \cite{Dill1997},
it can be trapped in a energy well during its visit in the conformational space. Even with the
advanced method like replica exchange molecular dynamics the three dimensional structure may be
trapped in a local minima for a considerable time. Due to various reasons it can also be misfolded
and trapped in a local energy minima. And one can mistakenly identify the misfolded state as the
native state (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

Calculation of free energy has been successfully attempted by a number of groups \cite{Ytreberg2006} - \cite{Zheng2008}. In
recent years methods like Refrence system method, Decativated Morphing \cite{Park2008}, Orthogonal Space Random
Work, Confinement Method have came out for calculation of free energy method. Some of the groups
relies on one of the great strength of free energy calculation, that it is a state function and thus
it is independent of path.

In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraint to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Then we have applied this method to larger proteins for the
first time. The free energy of conformation of pair of proteins with similar sequence but completely
different fold was calculated. Encouraged by the result, we applied the method of different targets
of CASP9 competition.

@@

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from
protein G, known as BHP \cite{Cecchini2009}. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations \cite{Cecchini2009},\cite{Krivov2004}  shows that bhp1 is the more favorable
configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value 1.7 kcal/mol, which
is in good agreement with the equilibrium simulations and with previous calculations using the
confinement method \cite{Cecchini2009} (see Supporting Information for further details).

@@

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. One of the  applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9 10, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimental structure (once it was available from the PDB). As
Table~\ref{table:casp_control} shows, in X of Y cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. Although
differentiating between experimentally determined structures and computer generated predictions is
not a stringent test of a scoring method---for example, examining the residue-residue packing can
easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
"sanity check".

@@

\begin{center}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $12.49 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $7.70 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $22.00 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.24 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $22.15 \pm 0.49$ \\ \hline 
\end{tabular}
\end{center}

@@

The fact such small changes in sequence can lead to such dramatic changes in structure is rather
remarkable. Accurately predicting the structural preferences of these structures presents a serious
challenge for computational methods \cite{Alexander2007}$-$\cite{Shortle20009}.

@@

The results of these calculations are presented in Figure~\ref{fig:orban}. The confinement method
identifies the correct structure for all six sequences. The free energy differences range from
around 2.9 to 5.7 kcal/mol, which is consistent with small changes in sequences and with estimates
made by Orban and co-workers \cite{He2008}, \cite{{Alexander2009}}.

\begin{figure}
\includegraphics[width=14cm,height=14cm]{orban.pdf}
\label{fig:orban}
\caption{The confinement method correctly predicts the structural preferences of six chameleon
sequences. (A) The six sequences used in this study. (B) Each sequence adopts either a Protein
G-like fold (denoted GB) or a three-helix bundle fold (denoted GA). The relative free energies of
the two folds are reported for each sequence.}
\end{figure}

@@

The native structure is a  67 residue NMR solved structure. We first reviewed all the five models
submitted by the group "BAKER-ROSETTASERVER", which was the best predictor group for this particular
target.  Among them we choose only the three structures, model 1, 3 and 5 to calculate the free
energy and subsequently identify the most stable structure among them. The rest of the two
structures are close to either of the selected structure and thus they are ommited from our
calculation. According to the GDT\_TS score, the model 1 was picked up correctly from the set of
structure generated . However, the order of model 3 and 5 was not correct.
It will be interesting to know whether confinement method can order the strcutures in terms
of free energy scale. The result is presented in Figure~\ref{fig:T0559}.

@@

Target T0560 is a 74 amino acid residue target from CASP9. However residues 3-66 were kept for final
result.  For this target we have compared the free energy of the model 2 and model 4 submitted by
the group "Splicer" along with the crystal structure.  The main difference between the model 2 and 4
is the orientation of the ....  The final result is presented in Figure~\ref{fig:T0560}.

@@

Target T0538 was a 54 residue target. The native structure was generated using NMR. The best model
that is closest to the native structure was generated by the group PconsR. However, their model 3
was better than their model 1. This structure was quite close to the NMR structure and had GDT\_TS
value of 96.23. We have calculated the free energy between this structure and the best structure
submitted by the group "Shell", "FOLDIT" and "BAKER-ROSETTASERVER". The calculated free energy value
is presented in Figure~\ref{fig:T0538}.

@@

Target T0540 was a 90 residue target. We have taken the best models from the group "LTB" which was
the best performing group for this target and also from the group Mufold and calculated the free
energy difference between these models with that of the native crystal structure.  As expected, the
crystal structure has the lowest free energy value, followed by Model 1 with GDT\_TS score of
$69.72$ and model 2 with GDT\_TS value of $53.789$. The final result is presented in
Figure~\ref{fig:T0540}.

@@

This is a 65 amino acid residue target with residues 6-63 were kept for final analysis. We have
calculated the free energy difference between all the five models submitted by the group
"MUFOLD-MD", which was one of the best performed group for this target.  The result is presented in
Figure~\ref{fig:T0531}.

@@

The best performing group was "Baker" for this target. We have carried out our calculation with the model first, second and fifth for the
final calculation. The rest two model had closer GDT\_TS value compared to the previously mentioned models and thus they were omitted from
furthe calculation. The result is presented in Figure~\ref{fig:T0605}. (This is a case where we can demand the confinement
method is better than GDT\_TS).

@@

@@

@@

@@

\begin{figure}
\begin{center}
\includegraphics[width=12cm,height=10cm]{T0540.pdf}
\end{center}
\caption{T0540}
\label{fig:T0540}
\end{figure}

@@

@@

@@

\begin{thebibliography}{99}

\bibitem{Dill1997}
Dill, K.A., and H.S. Chan.  From Levinthal to Pathways to Funnels:  The "New View" of Protein Folding Kinetics.  Nature Structural Biology 4: 10-19 (1997)

\bibitem{Ytreberg2006}
Ytreberg, F.; Zuckerman, D. Simple estimation of absolute free energies for biomolecules. J. Chem. Phys. 2006, 124, 104105.

\bibitem{Park2008}
Park, S.; Lau, A.; Roux, B. Computing conformational free energy by deactivated morphing. J. Chem. Phys. 2008, 129, 134102

\bibitem{Zheng2008}
Zheng, L.; Chen, M.; Yang, W. Random walk in orthogonal space to achieve efficient free-energy simulation of complex systems, Proc. Natl. Acad. Sci. 2008, 105 (51), 20227.

\bibitem{Tyka2006}
Tyka, M.; Clarke, A.; Sessions, R. An Efficient, Path-Independent Method for Free-Energy Calculations. J.Phys.Chem. B 2006, 110, 17212-17220.

\bibitem{Cecchini2009}
Cecchini, M., Krivov, S.V., Spichty, M., Karplus, M. Calculation of free-energy differences by confinement simulations. Application to peptide conformers. 
J. Phys. Chem. B 113, p. 9728-9740 (2009).

\bibitem{Krivov2004} 
Krivov, S.; Karplus, M. Hidden complexity of free energy surfaces for peptide (protein) folding Proc. Natl. Acad. Sci. U.S.A. 2004, 101, (41), 14766.

\bibitem{Alexander2007}
Alexander, P.A.; He, Y.; Chen, Y.; Orban, J. Bryan, P. The design and characterization of two proteins with $88 \%$ sequence identity but different 
structure and function. Proc. Natl. Acad. Sci. 2007, 104 (29), 11963-11968.

\bibitem{He2008}
He, Y.; Chen, Y.; Alexander, P.A.; Orban, J. NMR structures of two designed proteins with high sequence identity but different fold and function. Proc. Natl. Acad. Sci. 2008, 105 (38), 14412-14417.

\bibitem{Alexander2009}
Alexander, P.A.; He, Y.; Chen, Y.; Orban, J. Bryan, P. A minimal sequence code for switching protein structure and function. Proc. Natl. Acad. Sci. 2009, 106(50), 21149-21154.

\bibitem{Shortle20009}
Shortle, D. One sequence plus one mutation equals two folds. Proc. Natl. Acad. Sci. 2009, 106(50), 21011-21012.

\bibitem{Sheffler2009}
Sheffler, W.; Baker, D. RosettaHoles: Rapid assessment of protein core packing for structure prediction, refinement, design, and validation. Protein Science. 2009, 18(1), 229-239.

\end{thebibliography}
